---
title: "lab 4"
author: "Liz Gibbons"
date: "2026-04-17"
output: html_document
eval: false
categories:
  - lab
  - bios25328
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

::: {.callout-important}
## Before you start: download the lab data

Download the lab data from [Box](https://uchicago.box.com/s/snhot2x0272dzuuas4ipkgnrlxva2qga) and place it in a folder called `Lab-personal-genomes-project-data` inside your project directory. The folder must contain:

- `repgp-data.sqlite3`
- `repgp.fam`
- `repgp.chr1` … `repgp.chr22` (plink genotype files)
- `ukb_height.gz` (GWAS summary statistics)

Then set the `DATA` path in the R chunk below before running any code.
:::

```{r set-paths, eval=FALSE}
DATA <- "Lab-personal-genomes-project-data"  # relative to your project root; adjust if needed
```

## Goals and Introduction

In this lab we are going to use the PRSice software to take height GWAS results and predict height of the Personal Genome Project individuals using their genotype data.

### Goals

-   Download data from box
-   Download PRSice
-   Run exploratory analysis of data
-   Run PRSice
-   Plot observed versus predicted for height

## Setup

### load R packages and define paths

You will have to install all these packages first:

(r code block)
```{r, eval = F}
install.packages("tidyverse")
install.packages("RSQLite")
install.packages("glue")
```

(r code block)
```{r, eval = F}
suppressMessages(library(tidyverse))
suppressMessages(library(RSQLite))
suppressMessages(library(glue))
```

### Download data

-   download the data from [Box](https://uchicago.box.com/s/snhot2x0272dzuuas4ipkgnrlxva2qga) and place it in a directory called `Lab-personal-genomes-project-data`
-   download PRSice software from <https://choishingwan.github.io/PRSice/>

If you are on a Mac install the mac appropriate binary:

(bash code block)
```{bash, eval=FALSE}
# wget https://github.com/choishingwan/PRSice/releases/download/2.3.5/PRSice_mac.zip
```

If you are on Posit cloud do the linux binary:

(bash code block)
```{bash, eval=FALSE}
wget https://github.com/choishingwan/PRSice/releases/download/2.3.5/PRSice_linux.zip
```

Unzip the binary:
(bash code block)
```{bash, eval=FALSE}
unzip ./PRSice_linux.zip
```

move the executable to the directory where you keep your software (optional) \*\* the first time you run PRSice, it will install additional components make sure PRSice is executable \*\* chmod u+x PRSice

(bash code block)
```{bash, eval = F}
chmod +x ./PRSice_linux
```

Terminal Setup

(bash code block)
```{bash, eval = F}
export PRE="/cloud/project"          # change to your project root if not on Posit Cloud
export DATA="$PRE/Lab-personal-genomes-project-data"
export WORK=$DATA

export PRSice="$PRE"                 # directory containing PRSice.R and the binary
export PRSice_os="PRSice_linux"      # use PRSice_mac on a Mac
```

## Exploration of the Data

### open the database with phenotype information and list tables

(r code block)
```{r, eval = F}
dbname <- paste0(DATA,"/repgp-data.sqlite3") ##This is just to create the file path to the sqlite3 file 
## connect to db
db = RSQLite::dbConnect(SQLite(), dbname)
## list tables
dbListTables(db)
```

query database users table

(r code block)
```{r, eval = F}
dbListFields(db,"users")
```

(r code block)
```{r, eval = F}
query <- function(...) dbGetQuery(db, ...)
users = query("select * from users")
names(users)
```

look at distribution by count for each column

(r code block)
```{r, eval = F}
users %>% count(race)
```

(r code block)
```{r, eval = F}
users  %>% count(gender)
```

(r code block)
```{r, eval = F}
users %>% count(blood)
```

plot height distribution by gender

(r code block)
```{r, eval = F}
users %>% ggplot(aes(height,fill=gender)) + geom_density(alpha=0.6) + ggtitle("Height by gender - Missing gender, *, has bi-modal distr.") + theme_minimal(base_size = 15)
```

## run PRSice

### create phenotype data file

create phenotype data file as intersection of the plink formatted fam file and the users table from the database repgp-data.sqlite3

(r code block)
```{r, eval = F}
fam = read_tsv(glue::glue("{DATA}/repgp.fam"),col_names = FALSE)
```

(r code block)
```{r, eval = F}
names(fam)[1:2] = c("FID","IID")
fam <- fam %>% select(FID, IID) %>% inner_join(users %>% select(sample,height,weight,gender), by=c("IID"="sample")) 
write_tsv(fam,file=glue::glue("{DATA}/phenodata.txt"))
```

execute bash command to run PRSice

(bash code block)
```{bash, eval = F}
mkdir /cloud/project/Lab-personal-genomes-project-data/output

Rscript /cloud/project/PRSice.R --dir /cloud/project \
    --prsice /cloud/project/PRSice_linux \
    --base /cloud/project/Lab-personal-genomes-project-data/ukb_height.gz \
    --target cloud/project/Lab-personal-genomes-project-data/repgp.chr# \
    --snp variant_id \
    --A1 effect_allele \
    --A2 non_effect_allele \
    --stat effect_size \
    --beta \
    --pvalue pvalue \
    --pheno-file /cloud/project/Lab-personal-genomes-project-data/phenodata.txt \
    --pheno-col height \
    --bar-levels 5e-08,5e-07,5e-06,5e-05,5e-04,5e-03,5e-02,5e-01,1 \
    --fastscore \
    --binary-target F \
    --thread 2 \
    --out /cloud/project/Lab-personal-genomes-project-data/output/height_score_all
```

If you get duplicated SNP ID, follow the instructions

(bash code block)
```{bash, eval = F}
PRSice="/cloud/project/"
WORK="/cloud/project/Lab-personal-genomes-project-data/"

Rscript $PRSice/PRSice.R --dir $PRSice \
    --prsice $PRSice/PRSice_mac \
    --base $WORK/ukb_height.gz \
    --extract $WORK/output/height_score_all.valid \
    --target $WORK/repgp.chr# \
    --snp variant_id \
    --A1 effect_allele \
    --A2 non_effect_allele \
    --stat effect_size \
    --beta \
    --pvalue pvalue \
    --pheno-file $WORK/phenodata.txt \
    --pheno-col height \
    --bar-levels 5e-08,5e-07,5e-06,5e-05,5e-04,5e-03,5e-02,5e-01,1 \
    --fastscore \
    --binary-target F \
    --thread 2 \
    --out $WORK/output/height_score_all
```

This may take a couple minutes

plot observed vs predicted height

(r code block)
```{r, eval = F}
## we already have the phenotype data in the fam variable
predicted_height = read_delim(glue::glue("{DATA}/output/height_score_all.best"),delim=" ")
```

(r code block)
```{r, eval = F}
combined <- predicted_height %>% inner_join(fam,by=c("IID"="IID")) 
combined %>% ggplot(aes(PRS,height)) + geom_point() + theme_minimal(base_size = 15)
```

regress observed height with predicted height (PRS)

(r code block)
```{r, eval = F}
summary(lm(height ~ PRS,data=combined)) %>% coef() 
```

